---
title: "Heart Failure"
author: "Aaron Shikh, Stephen Shannon, Kevine Odira"
date: "2025-05-11"
output:
  ioslides_presentation:
    widescreen: true
  slidy_presentation: default
subtitle: "Predicting Heart Disease from Stress and ECG tests"
---

<style>
.remark-slide-content {
  background-color: #f7fcfc;
  color: #222;
  font-family: "Helvetica", "Arial", sans-serif;
  padding: 30px;
}

h1, h2 {
  color: #0b7285;
}

.title-slide h1 {
  color: #1864ab;
}

strong {
  color: #c92a2a;
}


</style>


```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(patchwork)
library(scales)
library(readr)
library(tidyr)

heart_data <- read.csv("heart.csv")

heart_data$HeartDisease <- as.factor(heart_data$HeartDisease)
heart_data$ChestPainType <- as.factor(heart_data$ChestPainType)
heart_data$Sex <- as.factor(heart_data$Sex)
```


## Research Question

**What factors contribute most to the presence of heart disease?**

We are particularly interested in:
- How the heart responds to stress (via ECG measurements)
- Whether specific test patterns help identify high-risk individuals



## Dataset Introduction

We use a publicly available dataset containing clinical measurements of patients undergoing cardiac screening.


## Variables
- **Demographics**
  - `Age`: Patient's age in years
  - `Sex`: Male or Female

- **Cardiac Symptoms**
  - `ChestPainType`: Type of chest pain (Typical Angina, Atypical, Non-anginal, Asymptomatic)
  - `ExerciseAngina`: Chest pain triggered by exercise (Yes/No)

- **Clinical Measurements**
  - `RestingBP`: Resting blood pressure (mm Hg)
  - `Cholesterol`: Serum cholesterol (mg/dL)
  - `FastingBS`: Binary indicator if fasting blood sugar > 120 mg/dL
  
- **Outcome Variable**
  - `HeartDisease`: 1 = has heart disease, 0 = no heart disease

## Variables relating to Stress Test & ECG Measurements
  - `MaxHR`: Maximum heart rate achieved
  - `ST_Slope`: Slope of the ST segment during peak exercise (Up/Flat/Down)
  - `Oldpeak`: ST depression during exercise (numeric)
  - `RestingECG`: Resting electrocardiogram results (Normal, ST abnormality, LVH)



## Key Predictors: RestingECG, ST_Slope, OldPeak

`RestingECG`: describes the heart's electrical activity at rest:
  - **Normal**: No abnormalities in ST segment or T waves.
  - **ST**: ST-T wave abnormality — includes T wave inversions or ST elevation/depression.
  - **LVH**: Left Ventricular Hypertrophy — signals enlarged heart muscle, often from high blood pressure.

`ST_Slope`: describes the shape of the ST segment during peak exercise:
  - **Up**: ST segment increases — this is the normal, healthy response.
  - **Flat**: No change in ST segment — may indicate ischemia.
  - **Down**: ST segment decreases — strongest indicator of potential heart disease.

`Oldpeak`: represents the amount of ST segment depression measured during exercise.  
  - A higher value indicates potential **ischemia** (oxygen deprivation in the heart).
  - Measured in millimeters (mm)



## Distribution of Oldpeak by ST_Slope

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
heart_data$ST_Slope <- factor(heart_data$ST_Slope, levels = c("Up", "Flat", "Down"))

ggplot(heart_data, aes(x = ST_Slope, y = Oldpeak, fill = ST_Slope)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.size = 0.8, color = "black") +
  labs(title = "Oldpeak by ST_Slope",
       x = "ST Slope",
       y = "Oldpeak (ST Depression)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Box plots

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
box_plot1 <- ggplot(heart_data, aes(x = Sex, y = MaxHR, fill = HeartDisease)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "MaxHR vs Sex", x = "Sex", y = "MaxHR", fill = "Heart Disease") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
        plot.title = element_text(hjust = 0.5, face = "bold"))

box_plot2 <- ggplot(heart_data, aes(x = Sex, y = Age, fill = HeartDisease)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Age vs Sex", x = "Sex", y = "Age", fill = "Heart Disease") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
        plot.title = element_text(hjust = 0.5, face = "bold"))

(box_plot1 | box_plot2) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "right")


```

## Bar Plot

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
grouped_bar_plot<- ggplot(heart_data, aes(x = ChestPainType, fill = HeartDisease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent) +
  labs(title = "Heart Disease by Chest Pain Type",
       x = "Chest Pain Type", y = "Proportion", fill = "Heart Disease") +
theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.7),
        plot.title = element_text(hjust = 0.5, face = "bold"))
print(grouped_bar_plot)

```

## Scatterplot

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
plot_ly(
  data = heart_data,
  x = ~MaxHR,
  y = ~Age,
  z = ~Cholesterol,
  color = ~HeartDisease,
  colors = c('blue', 'red'),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 4, opacity = 0.7, line = list(width = 0.5, color = 'gray'))
) %>%
  layout(
    title = "MaxHR vs Age vs Cholesterol",
    scene = list(
      xaxis = list(title = "Max Heart Rate"),
      yaxis = list(title = "Age"),
      zaxis = list(title = "Cholesterol (mg/dl)")
    ),
    legend = list(title = list(text = "<b>Heart Disease</b>"))
  )
```

## Heatmap

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
heart_data$HeartDisease <- as.numeric(as.character(heart_data$HeartDisease))
old_peak_binned <- heart_data %>%
  mutate(OldpeakBin = cut(Oldpeak,
    breaks = c(-Inf, 0, 1, 2, 3, Inf),
    labels = c("0", "0–1", "1–2", "2-3", ">3"),
    right = TRUE, include.lowest = TRUE))


heatmap_df <- old_peak_binned %>%
  group_by(RestingECG, OldpeakBin) %>%
  summarise(HeartDiseaseRate = mean(HeartDisease), .groups = "drop") %>%
  pivot_wider(names_from = OldpeakBin, values_from = HeartDiseaseRate)

z_matrix <- as.matrix(heatmap_df[,-1])
rownames(z_matrix) <- heatmap_df$RestingECG

plot_ly(
  x = colnames(z_matrix),
  y = rownames(z_matrix),
  z = z_matrix,
  type = "heatmap",
  colorscale = "Reds",
  hovertemplate = "RestingECG: %{y}<br>Oldpeak: %{x}<br>Rate: %{z:.2f}<extra></extra>"
) %>%
  layout(
    title = "Heart Disease Rate by RestingECG and Oldpeak",
    xaxis = list(title = "Oldpeak"),
    yaxis = list(title = "Resting ECG")
  )

```






