---
title: "Presentation_slides"
author: "Group 8, Kevine, A"
date: "2025-05-11"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE, eval = FALSE}

# 	* Research question: What patterns in clinical measurements and symptoms are associated with heart disease,
#  and how do they vary across demographic groups?
# 
# 	* Why it matters:can help reveal which demographics (age, sex) tend to experience
#  more severe or different symptom profiles /early warning signs in specific groups.
# 
# 	* What datasets are we using
# 		* https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction/data
# 
# 
# 	* Potential questions to explore: 
# 		* How does Age and Gender relate to heart disease?
# 		* What clinical markers are associated with heart disease?
# 
# 
# 	* Interesting visualizations
# 		* Corrplot among clinical markers -- strong association between certain markers?
# 		* Grouped boxplot of RestingECG Categories and proportion w/ heart disease
# 		* 3d plot between MaxHR and Oldpeak, see the joint distribution
# 		* Potential shiny plotly app: Select different binary factors and see charts of how risk
# 	 of heart disease changes based on selection, or how distributions of other variables change
# 		** Needs some type of logistic reg model **
# 
# 	* Limitations:
# 		* Missing other potentially important factors like weight, excercise level, diet type, etc.





```

## Slide with Plot

```{r pressure}
plot(pressure)
```

